---
title: "R Warm-up for Visualisation"
author: "Priit Adler, Nurlan Kerimov"
date: "2026-04-02"
output: html_document
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(tibble)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

cars <- mtcars %>%
  rownames_to_column(var = "car") %>%   # move row names (car models) into a real column
  mutate(
    cyl = factor(cyl),                  # treat cylinder count as a category, not a number
    gear = factor(gear),                # same for number of gears
    am = factor(am, levels = c(0, 1), labels = c("Automatic", "Manual"))  # give 0/1 readable labels
  )
```

## Why this document?

This warm-up is a short prerequisite companion for the ggplot2 session. Use it
to refresh the small data operations that often happen just before plotting.

## Course preconditions

Before coming to the visualisation course, you should be comfortable with:

-   basic R and RStudio navigation
-   running code chunks in R Markdown
-   the idea of tidy data
-   the difference between categorical and continuous variables
-   factors as stored categories

If these ideas feel rusty, the quick refresher below will help.

## Quick refresher

### Tidy data

In tidy data, every row is one observation and every column is one variable. The
`cars` dataset is tidy: each row is one car, and each column describes a
property of that car.

### Factors

A factor is how R stores categorical variables — values that belong to a fixed
set of categories rather than being free-form numbers or text. For example, the
`cyl` column holds only three possible values:

```{r}
levels(cars$cyl)
```

Factors matter for plotting because they control legend ordering, colour
assignment, and axis grouping.

### The pipe (`%>%`)

We use the pipe operator `%>%` to chain operations — read it as "and then." For
example, `cars %>% select(mpg)` means "take the `cars` table, and then select
the `mpg` column." Every exercise below uses this pattern.

## Meet the dataset

In this warm-up we use the built-in `mtcars` dataset for every example. We've
made a few adjustments to the data to make it more user-friendly. Now it is
called `cars`:

```{r}
glimpse(cars)
```

The main variables we will work with are:

| Column | Type        | Description                          |
|--------|-------------|--------------------------------------|
| `car`  | text        | car name (added during setup)        |
| `mpg`  | continuous  | miles per gallon (fuel efficiency)   |
| `hp`   | continuous  | gross horsepower                     |
| `wt`   | continuous  | weight in 1000 lb                    |
| `disp` | continuous  | engine displacement in cubic inches  |
| `cyl`  | categorical | number of cylinders (4, 6, or 8)     |
| `gear` | categorical | number of forward gears (3, 4, or 5) |
| `am`   | categorical | transmission (Automatic or Manual)   |

## Exercise 1: `select()`

Pick a small set of columns that would be useful in a fuel-efficiency summary:
`car`, `mpg`, `hp`, `wt`, and `am`.

```{r}
cars %>%
  select(car, mpg, hp, wt, am) %>%  # keep only these five columns, drop the rest
  head()                             # show the first 6 rows (default)
```

Expected result: a 6-row table with only the requested five columns.

Note: remember that `select()` changes columns, not rows.

## Exercise 2: `filter()`

Find manual cars with more than 25 miles per gallon.

```{r}
manual_high_mpg <- cars %>%
  filter(am == "Manual", mpg > 25)  # both conditions must be true: manual AND efficient

manual_high_mpg
```

Expected result: a short table of efficient manual cars such as the Fiat 128,
Honda Civic, and Lotus Europa.

Note: filtering is useful for plotting too — for example, you might want to
show only 4-cylinder cars or focus on one treatment group.

## Exercise 3: `arrange()`

Sort the data from highest to lowest fuel efficiency. If two cars have similar
fuel efficiency, place the lighter car first.

```{r}
cars %>%
  arrange(desc(mpg), wt) %>%  # desc() reverses order for mpg; wt stays ascending as tiebreaker
  head(10)                     # show only the top 10 rows
```

Expected result: the most fuel-efficient cars appear first, with light cars
breaking ties.

Note: you can mix `desc()` with regular ascending ordering inside the same
`arrange()` call.

## Exercise 4: `mutate()`

Create two new variables:

-   `power_to_weight` = `hp / wt`
-   `mpg_band` with values `"low"`, `"medium"`, and `"high"`

Use this rule for `mpg_band`:

-   `"high"` if `mpg >= 25`
-   `"medium"` if `mpg >= 18` and `< 25`
-   `"low"` otherwise

`case_when()` checks conditions in order and returns the value for the first one
that is true — like a series of if/else statements.

```{r}
cars_augmented <- cars %>%
  mutate(
    power_to_weight = hp / wt,        # derived ratio: higher means more powerful for its weight
    mpg_band = case_when(             # case_when works like a multi-condition if/else
      mpg >= 25 ~ "high",             # first matching condition wins
      mpg >= 18 ~ "medium",           # only reached if mpg < 25
      TRUE ~ "low"                    # TRUE is the catch-all fallback (like else)
    )
  )

cars_augmented %>%
  select(car, mpg, hp, wt, power_to_weight, mpg_band) %>%  # narrow down to relevant columns
  head(10)
```

Expected result: the output includes one continuous derived variable and one
categorical variable.

Note: creating categories like `mpg_band` is a natural bridge to plotting —
you will often use derived variables like this for colour or grouping.

## Exercise 5: `group_by()` + `summarise()`

Create a summary table by number of cylinders. Include:

-   number of cars
-   mean `mpg`
-   mean `hp`

```{r}
cars %>%
  group_by(cyl) %>%          # split into groups: one per cylinder count (4, 6, 8)
  summarise(
    n_cars = n(),             # n() counts rows in each group
    mean_mpg = mean(mpg),
    mean_hp = mean(hp)
  )
```

Expected result: one row each for 4, 6, and 8 cylinders.

Note: `count(cyl)` is a handy shortcut for
`group_by(cyl) %>% summarise(n = n())` when you only need counts.

Note: summarised tables like this are perfect raw material for bar plots, line
plots, and heatmaps in the visualisation course.

## Exercise 6: grouped summary with two variables

Create a summary table by transmission and cylinders. This is the kind of table
that later becomes a bar plot, box plot, or heatmap.

Include:

-   `am`
-   `cyl`
-   mean `mpg`
-   mean `wt`

```{r}
summary_by_transmission <- cars %>%
  group_by(am, cyl) %>%          # group by two variables: one row per unique am+cyl combination
  summarise(
    mean_mpg = mean(mpg),
    mean_wt = mean(wt)
  )

summary_by_transmission          # print the saved result
```

Expected result: a compact summary table with one row per transmission and
cylinder combination.

Note: this exact table shape is what many plotting geoms expect as input.

## Exercise 7: creating a factor

The `vs` column is stored as a number (0 or 1) but it actually represents a
category: engine shape. Convert it into a factor with readable labels.

0: V-shaped
1: Straight

```{r}
cars %>%
  mutate(
    vs = factor(vs, levels = c(0, 1), labels = c("V-shaped", "Straight"))
  ) %>%
  select(car, vs, cyl, mpg) %>%
  head()
```

Expected result: the `vs` column now shows `"V-shaped"` or `"Straight"` instead
of 0 or 1.

Note: in the visualisation course, factor conversion like this controls how
categories appear in legends, axes, and colour assignments.

## Checkpoint

These small verbs show up again and again when building plots:

-   `select()` helps you focus on the variables you want to show
-   `filter()` limits the data to a relevant subset
-   `arrange()` helps you inspect patterns before plotting
-   `mutate()` creates variables for labels, colours, groups, or facets
-   `group_by()` and `summarise()` create summary tables for bars, lines, and
    heatmaps
-   `factor()` turns numbers or text into proper categories with labels

In the main visualisation course, we will assume you can read and modify tables
at about this level.
